Non-convergence in iterative imputation

### 2025-05-08 Manuscript/manuscript.qmd
cross-ref figures

@@

- can we diagnose non-convergence quantitatively?
- ~~which parameter should we track?~~
- how many iterations are needed before converge to a stationary distribution is reached?
- how many iterations are needed before valid inferences are reached?

@@

[TODO: add *approximate* convergence as the most converged state that the algorithm can obtain?]

@@

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $\mathbf{Y}_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $\mathbf{Y}_{\mathrm{obs}}$ to create $m$ completed datasets. On each of these datasets, the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

```{r diagram, echo=FALSE, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}

@@

**Algorithm 1: MICE algorithm**

*Initialization phase ($t = 0$)*

@@

*Iteration phase ($t = 1,...,\tau$)*

@@

<!-- \begin{enumerate} -->
<!--    \item First level item -->
<!--    \item First level item -->
<!--    \begin{enumerate} -->
<!--      \item Second level item -->
<!--      \item Second level item -->
<!--      \begin{enumerate} -->
<!--        \item Third level item -->
<!--        \item Third level item -->
<!--      \end{enumerate} -->
<!--    \end{enumerate} -->
<!--  \end{enumerate} -->

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13]. With a recently proposed adaptation, $\widehat{R}$ may also serve to diagnose non-stationarity [@vehtari2021]. We will evaluate the appropriateness of $\widehat{R}$ and $AC$. Other methods are outside the scope of this study, e.g. because they assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.

@@

We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; @R). We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI]. The number of simulation repetitions $n_{\mathrm{sim}} = 2000$.

@@

**Algorithm 2: simulation set-up (pseudo-code)**
```
for (each simulation run) {
  simulate complete data;
  for (each missingness condition) {
    create missingness;
    for (each imputation model iteration) {
      impute missingness;
      estimate quantities of scientific interest;
      apply performance measures to the estimates;
      compute non-convergence diagnostics
} } }
```

@@

\begin{table}[]
\begin{tabular}{lcr}
\hline
\multicolumn{3}{l}{Missingness conditions} \\ \hline
missing data mechanism &  & proportion of incomplete cases  \\
                       & $\times$  &                         \\
MCAR, MAR              &   & 25\%, 50\%, 75\%                \\ \hline
\end{tabular}
\end{table}

@@

[TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018?].

@@

@@

We use two non-convergence identifiers---autocorrelation and $\widehat{R}$---to diagnose non-convergence in the imputation models of the four incomplete variables. For each variable we apply the two identifiers on two parameters---chain means and chain variances---and four variables, resulting in 16 sets of identifier-parameter-variable pairs.

- non-convergence in iterative algorithms is diagnosed using an identifier and a parameter

- the parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means)

- the identifier is a calculation of some sort that quantifies non-convergence

- identifiers are historically focused on either non-mixing between chains or non-stationarity within chains

- the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well

- just to be sure, we also use autocorrelation to quantify trending within chains

- we apply these identifiers to the two parameters that we typically evaluate after imputation using visual inspection: chain means and chain variances

@@

[TODO: add tables, with MC SEs]

@@

```{r bias}
plot_results(results, "bias", 0, "MCAR")
```

- more missingness = more extreme bias, but also steeper decrease over iterations

```{r cov}
plot_results(results, "cov", 0.95, "MCAR")
```

- no clear trend

```{r ciw}
plot_results(results, "ciw", NULL, "MCAR")
```

- more missingness = wider CIs, with steeper decrease in CIW, but not stababiling at same value between missingness conditions.

In Figure \@ref(fig:perf) we show the performance measures: bias in the regression estimate (panel A), the empirical coverage rate of the regression estimate (panel C) and the average confidence interval width of this estimate (panel D).

We see that within a few iterations the bias in the regression estimate approaches zero (see \@ref(fig:perf)A). When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates [regression coefficient is underestimated because there is less info to estimate the relation??].

Nominal coverage is quickly reached (see \@ref(fig:perf)C). After just three iterations, the coverage rates are non-improving in every missingness condition [but MNAR with 5% incomplete cases does not reach nominal coverage --\> due to bias in the estimate in combination with very narrow CI (see CIW!)].

The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see \@ref(fig:perf)D). Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$.

Figure \@ref(fig:ac) shows the autocorrelations in the chain means (panel A) and chain variances (panel B).

```{r ac_mean}
plot_results(results, "ac_mean", 0, "MCAR")
```

```{r ac_sd}
plot_results(results, "ac_sd", 0, "MCAR")
```

Autocorrelation in the chain means decreases rapidly in the first few iterations (see \@ref(fig:ac)A). The decrease is substantive until $n_{it} \geq 6$. This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity. These results hold irrespective of the missingness condition.
Autocorrelation in the chain variances show us something similar (see \@ref(fig:ac)B). The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $n_{it} \geq 10$. We do not observe a systematic difference between missingness conditions here either.

```{r psrf_mean}
plot_results(results, "psrf_mean", 1, "MCAR")
```

```{r psrf_sd}
plot_results(results, "psrf_sd", 1, "MCAR")
```

Figure \@ref(fig:rh) shows the potential scale reduction factor in the chain means (panel A) and chain variances (panel B).

We observe that $\widehat{R}$-values of the chain means generally decrease as a function of the number of iterations (see \@ref(fig:rh)A). An exception to this observation is a steep increase in iterations $3 \leq n_{it} \leq 5$ [TODO: interpret?? due to initialization or is there really more mixing initially??]. After the first couple of iterations, the mixing between chain means generally improves until $n_{it} \geq 30$ to $40$. There is no apparent differentiation between the missingness conditions. 
The mixing between chain variances mimics the mixing between chain means almost perfectly (see \@ref(fig:rh)B). Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $n_{it} \geq 30$. 
[The rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]

```{r, cache=TRUE, eval=FALSE, fig.height=9}
performance |>
  filter(method == "MICE") |>
ggplot(aes(x = .it, y = psrf_mean)) +
  geom_point(aes(group = .sim), shape = 1, alpha = 0.05) +
  geom_smooth(method = "loess") +
  facet_grid(term ~ mech + prop) +
  labs(
    title = "Potential Scale reduction Factor(PSRF) applied to chain means",
    x = "Iteration number",
    y = "PSRF"
  ) +
  theme_classic() 
```

@@

@@

Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the non-convergence identifiers. Of course, it never hurts to iterate longer, but such calculations hardly bring added value.

- Convergence diagnostics keep improving substantially until n_it = 20-30

- Performance measures do not improve after n_it = 9

- [methodological explanation is that rhat and ac have a lag (few it to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]

@@

From the traceplot of the chain means (see \@ref(fig:case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.

This figure (\@ref(fig:case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.

The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see \@ref(fig:case)C).

According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see \@ref(fig:case)D).

>>>>

<!-- ## Aims -->

<!-- - can we diagnose non-convergence quantitatively? -->
<!-- - ~~which parameter should we track?~~ -->
<!-- - how many iterations are needed before converge to a stationary distribution is reached? -->
<!-- - how many iterations are needed before valid inferences are reached? -->

@@

@@

<!-- *approximate* convergence as the most converged state that the algorithm can obtain (gelman, 2013?) -->

@@

Figure @fig-diagram provides an overview of the steps involved with MI. The missing part $\mathbf{Y}_{\mathrm{mis}}$ of an incomplete dataset is imputed $m$ times. This creates $m$ sets of imputed data $\dot \mathbf{Y}_{\mathrm{imp}, \ell}$, where $\ell = 1, 2, ..., m$. The imputed data is then combined with the observed data $\mathbf{Y}_{\mathrm{obs}}$ to create $m$ completed datasets. On each of these datasets, the analysis of scientific interest is performed to estimate $Q$: the quantity of scientific interest (e.g., a regression coefficient). Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}_\ell$-values are obtained. Finally, the $\hat{Q}_\ell$-values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence-valid estimate of the true---but unobserved---scientific estimand $Q$ [@rubin96].
<!-- "A scientific estimand $Q$ is a quantity of scientific interest that we can calculate if we would observe the entire population" [@buur18, par 2.3.1] -->

```{r fig-diagram, echo=FALSE, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation---from an incomplete dataset, to $m=3$ multiply imputed datasets, to $m=3$ estimated quantities of scientific interest $\\hat{Q}$, to a single pooled estimate $\\bar{Q}$."}

@@

::: {#nte-algorithm .callout-note}

## Algorithm 1: MICE algorithm

**Initialization phase ($t = 0$)**

@@

**Iteration phase ($t = 1,...,\tau$)**

@@

:::
<!-- \begin{enumerate} -->
<!--    \item First level item -->
<!--    \item First level item -->
<!--    \begin{enumerate} -->
<!--      \item Second level item -->
<!--      \item Second level item -->
<!--      \begin{enumerate} -->
<!--        \item Third level item -->
<!--        \item Third level item -->
<!--      \end{enumerate} -->
<!--    \end{enumerate} -->
<!--  \end{enumerate} -->

@@

<!-- - identifiers are historically focused on either non-mixing between chains or non-stationarity within chains -->

<!-- - the popular non-mixing identifier rhat has recently been updated by Vehtari et al, and should now work for non-stationarity as well -->

<!-- - just to be sure, we also use autocorrelation to quantify trending within chains -->

<!-- - we apply these identifiers to the two parameters that we typically evaluate after imputation using visual inspection: chain means and chain variances -->

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. Non-convergence is diagnosed using an identifier and a parameter. The parameter can be any statistic that we track across iterations, for example the average imputed value per imputation (i.e., chain means). The identifier is a calculation of some sort that quantifies non-convergence in the scalar. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ to evaluate mixing ['Gelman-Rubin statistic'; @gelm92], and autocorrelation to diagnose trending [$AC$; @scha97; @gelm13]. With a recently proposed adaptation, $\widehat{R}$ may also serve to diagnose non-stationarity [@vehtari2021]. We will evaluate the appropriateness of $\widehat{R}$ and $AC$. Other methods are outside the scope of this study, e.g. because they assume that values within chains represent independent samples, whereas the MICE algorithm only uses the final iteration to produce imputations.

@@

We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.4.2; @R). We provide a summary of the simulation set-up in Algorithm 2; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint) [TODO: make this a Zenodo DOI]. The number of simulation repetitions $n_{\mathrm{sim}} = 2000$.

@@

::: {#nte-simulation .callout-note}

## Algorithm 2: simulation set-up (pseudo-code)

```
for (each simulation run) {
  simulate complete data;
  for (each missingness condition) {
    create missingness;
    for (each imputation model iteration) {
      impute missingness;
      estimate quantities of scientific interest;
      apply performance measures to the estimates;
      compute non-convergence diagnostics
} } }
```
:::

@@

\begin{table}[]
\caption{Missingness scenarios}
\begin{tabular}{lcr}
\hline
\multicolumn{3}{l}{Missingness conditions} \\ \hline
missing data mechanism &  & proportion of incomplete cases  \\
                       & $\times$  &                         \\
MCAR, MAR              &   & 25\%, 50\%, 75\%                \\ \hline
\end{tabular}
\end{table}

@@

<!-- [TODO: add left-tailed M(N)AR mechanism, md pattern, and maybe refer to ampute paper or "Dance of..." by Schouten 2018?]. -->

@@

We use two non-convergence identifiers---autocorrelation and $\widehat{R}$---to diagnose non-convergence in the imputation algorithm. These two identifiers are applied to two scalar summaries of the state-space of the algorithm---chain means and chain variances. We compute the two identifiers on four different variables: the outcome variable $Y$, and the three predictors $X_1$, $X_2$, and $X_3$. In total, we apply the two identifiers on two parameters and four variables, resulting in 16 sets of identifier-parameter-variable pairs.

@@

@@

Our results show the non-convergence metrics and the performance of the MICE algorithm under different missingness conditions at a varying number of iterations. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias and coverage, $X_3$. Full results are available from [TODO: add repo DOI]. [TODO: add tables, with MC SEs, to appendix.]

### AC

Figures @fig-ac_mean and @fig-ac_sd show the autocorrelations in the chain means and chain variances, respectively.

```{r fig-ac_mean}
plot_results(results, "ac_mean", 0, "MCAR")
```

Autocorrelation in the chain means decreases rapidly in the first few iterations (see @fig-ac_mean). The decrease is substantive until $n_{it} \geq 6$. This means that there is some initial trending within chains, but the average imputed value quickly reaches stationarity. These results hold irrespective of the missingness condition.

```{r fig-ac_sd}
plot_results(results, "ac_sd", 0, "MCAR")
```

Autocorrelation in the chain variances show us something similar (see @fig-ac_sd). The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $n_{it} \geq 10$. We do not observe a systematic difference between missingness conditions here either.

### $\widehat{R}$

Figure @fig-psrf_mean and @fig-psrf_sd show the potential scale reduction factor in the chain means and chain variances respectively.

```{r fig-psrf_mean}
plot_results(results, "psrf_mean", 1, "MCAR")
```

We observe that $\widehat{R}$-values of the chain means generally decrease as a function of the number of iterations (see @fig-psrf_mean). An exception to this observation is a steep increase in iterations $3 \leq n_{it} \leq 5$ [TODO: interpret?? due to initialization or is there really more mixing initially??]. After the first couple of iterations, the mixing between chain means generally improves until $n_{it} \geq 30$ to $40$. There is no apparent differentiation between the missingness conditions.

```{r fig-psrf_sd}
plot_results(results, "psrf_sd", 1, "MCAR")
```

The mixing between chain variances mimics the mixing between chain means almost perfectly (see @fig-psrf_sd. Irrespective of the missingness condition, the $\widehat{R}$-values taper off around $n_{it} \geq 30$.

[TODO: write about: The rhat plots all show some initialization before the fifth iteration: is rhat useful before that??]

@@

In Figures @fig-bias - @fig-ciw we show the performance measures: bias in the regression estimate, the empirical coverage rate of the regression estimate and the average confidence interval width of this estimate.

```{r fig-bias}
plot_results(results, "bias", 0, "MCAR")
```

We see that within a few iterations the bias in the regression estimate approaches zero (see @fig-bias). When $n_{it} \geq 6$, even the worst-performing conditions (e.g., with a proportion of incomplete cases of 75%) produce stable, non-improving estimates.  [TODO: check if regression coefficients are first underestimated because there is less info to estimate the relation??] [TODO: write about: more missingness = more extreme bias, but also steeper decrease over iterations].

```{r fig-cov}
plot_results(results, "cov", 0.95, "MCAR")
```

Nominal coverage is quickly reached (see @fig-cov). After just three iterations, the coverage rates are non-improving in every missingness condition. 
<!-- no clear trend across conditions-->

```{r fig-ciw}
plot_results(results, "ciw", NULL, "MCAR")
```

The average confidence interval width decreases quickly with every added iteration until a stable plateau is reached (see @fig-ciw). Depending on the proportion of incomplete cases this takes up-to $n_{it} \geq 9$. [TODO: write about: more missingness = wider CIs, with steeper decrease in CIW, but not stababiling at same value between missingness conditions.]

<!-- [but MNAR with 5% incomplete cases does not reach nominal coverage; due to bias in the estimate in combination with very narrow CI (see CIW!)]. -->

@@

## Summary

Overall, the methods to diagnose non-convergence perform as expected: they indicate more signs of non-convergence in conditions with worse performance in terms of bias and confidence-validity. We observe approximately unbiased estimates after at most seven iterations (for any $\pi_{\mathrm{inc}}$ considered). This implies that the algorithm produces stable, non-improving estimates when $\tau\geq7$. The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to a $\widehat{R}$ threshold of $\approx 1.2$.

<!-- The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences. The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general. A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $\tau$, approximate stationarity may be concluded. -->

@@

Identifying non-convergence may be a crucial aspect of drawing valid statistical inferences from incomplete data. In this simulation study, we have shown that non-convergence in iterative imputation algorithms does indeed go hand in hand with biased, invalid estimates. Our study , however, found that inferential validity was achieved after five to ten iterations, much earlier than indicated by non-convergence identifiers. While  convergence diagnostics kept improving substantially until $\tau \approx 20-30$, performance measures did not improve after $\tau = 9$. Of course, it never hurts to iterate longer, but such calculations hardly brought added value.

The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached. Under the current specifications, approximately unbiased, confidence-valid estimates were obtained after a maximum of nine iterations. Continued iterations beyond $\tau = 9$ did not yield better estimates.
In contrast, $\widehat{R}$-values decreased substantially until 20 to 30 iterations, which implies that mixing in the algorithm still improved with each additional iteration. $AC$-values only improved until about six iterations, suggesting minimal improvement in trending was obtained beyond $\tau=6$ in the current simulation set-up.

[TODO: write about: methodological explanation is that rhat and ac have a lag (too few iterations to inform your statistic) --\> will always indicate convergence slower than inferential validity is reached]

@@

From the traceplot of the chain means (see @fig-case)A) it seems that mixing improves up-to 10 iterations, while trending is only apparent in the first three iterations.

This figure (@fig-case)B) shows that 7 iterations are required before the $\widehat{R}$-values of the chain means drop below the threshold for non-convergence.

The $\widehat{R}$-values for the scientific estimate reaches the threshold much sooner, when $n_{it}=14$ (see @fig-case)C).

According to the $\widehat{R}$-values with $\lambda$ as parameter, at least 15 iterations are required (see @fig-case)D).
